---
title: "Preprocessing: Where is Jiji?"
subtitle: Investigating Cat Movement, Space Use, and Environmental Influences
author: Katharina Genucchi and Selina Lepori
format:
  html:
    theme:
      dark: darkly
editor: visual
---

## Load Libraries

clear environment

```{r}
rm(list = ls())
```

```{r}
#| warning: false

library(sf)
library(lubridate)
library(adehabitatHR)
library(suncalc)
library(tidyverse)
library(tmap)
library(adehabitatHR)
library(quarto)

# source("functions_dist_secs.R")
# source("functions_territory_estimation.R")
# source("function_group.R")
```

## Functions

The following function for KDE was generated with the help of [ChatGPT](https://chatgpt.com/share/680fa817-0784-8012-a295-775907709f60){target="_blank"}

```{r}
# function calculating difference between two data points by second
difftime_secs <- function(later, now){
  as.numeric(difftime(later, now, units = "secs"))
}

# functions for calculating distance by element and time difference.
distance_by_element <- function(later, now) {
  as.numeric(
    st_distance(later, now, by_element = TRUE)
  )
}

# function for calculating speed  
function_speed <- function(data, datetime, geometry) {
  data <- data |>
    arrange(datetime) |> # make sure order is correct
    mutate(timelag = difftime_secs(later=lead(datetime), now=datetime),
           steplength = distance_by_element(later=lead(geometry), now=geometry),
           speed_ms = (steplength/timelag), #Calculate speed in m/s  
           speed_kmh = (steplength/timelag)*3.6 # Calculate speed in km/h
    )
  return(data)
}

# function to calculate mcp per month
# summarise() unions all points into a multipoint object.
function_mcp <- function(data, ID) {
  
  data_multipoint <- data |>
    group_by(month) |> 
    summarise()
  
  # calculate the convex hull
  data_mcp <- st_convex_hull(data_multipoint)
  
  data_mcp <- data_mcp |>
    mutate(area_ha = as.numeric(st_area(geometry))/10000,
           ID = ID,
           method = "mcp")
  
  return(data_mcp)
}

# function for kernel density estimation
function_kde <- function(data, ID) {
  
  separation <- unique(data$month)
  
  # initialise list to hold sf objects
  kde_list <- lapply(separation, function(m) {
    
    data_sep <- data[data$month == m, ]
    
    
    # convert to SpatialPoints
    sp_points <- as(data_sep, "Spatial")
    
    # estimate utilisation distribution using kernel method
    # probabilistic estimate of space use based on the input point locations
    # This is a surface showing the probability of the animal being in each location.
    kud <- kernelUD(sp_points)
    
    # get 95% territory as SpatialPolygons
    # takes this utilisation distribution and extracts the home range polygon that contains 95% of the estimated utilisation, i.e., the area where the animal is estimated to spend 95% of its time. This is a common way to define an animal’s home range.
    terr_kde <- getverticeshr(kud, percent=95)
    
    # ensure the correct CRS and make an sf out of st
    terr_sf_kde <- st_transform(st_as_sf(terr_kde), st_crs(data))
    
    # Reset rownames
    rownames(terr_sf_kde) <- NULL
    
    # Add month column
    terr_sf_kde$month <- m
    
    # keep geometry only
    terr_sf_kde <- terr_sf_kde["geometry"]
    
    # Add month column
    terr_sf_kde$month <- m
    
    return(terr_sf_kde)
  })
  # Combine all months into one sf object
  all_kde <- do.call(rbind, kde_list)
  
  all_kde <- all_kde |>
    mutate(area_ha = as.numeric(st_area(geometry))/10000,
           ID = ID,
           method = "kde")
  
  return(all_kde)
}

# function for grouping by Ground and calculate proportion of land cover according to territory estimation method
function_group <- function(data, data_land_cover, method_area, method_area_perc, buffer=FALSE) {
  
  data_new <- st_intersection(data, data_land_cover)
  
  if (buffer==TRUE) {
    
    data_new <- data_new |>
      group_by(Ground) |>
      summarise(area = sum(st_area(geometry))) |> 
      mutate(area = as.numeric(area)/10000,
             area_perc=round((area/sum(area)*100), digits=2))
  }
  
  else {
    data_new <- data_new |>
      group_by(month, Ground) |>
      summarise(area = sum(st_area(geometry))) |> 
      mutate(area = as.numeric(area)/10000,
             area_perc=round((area/sum(area)*100), digits=2))
    data_new <- subset(data_new, select=-c(geometry))
  }
  
  names(data_new)[names(data_new)=="area"] <- method_area
  names(data_new)[names(data_new)=="area_perc"] <- method_area_perc
  
  data_new <- st_drop_geometry(data_new)
  
  return(data_new)
}
```

## Read in and Tidy Data

### Land Use Data

#### Read in

```{r}
# read in geopackage containing lang cover
land_cover <- st_read("data/land_cover.gpkg", layer = "layer")

# rename geom to geometry, ensure correct geometry column
names(land_cover)[names(land_cover)=="geom"] <- "geometry"
st_geometry(land_cover) <- "geometry"

# select only necessary columns
land_cover <- subset(land_cover, select=c(Art_TXT, geometry))

# geometry in metres, not degrees / ensure correct CRS
land_cover <- st_transform(land_cover, crs=2056)
```

#### Group Land Use Data

```{r}
land_cover$Ground <- NA

land_cover$Ground[grepl("Gebaeude", land_cover$Art_TXT)] <- "Building"

land_cover$Ground[grepl("humusiert", land_cover$Art_TXT)] <- "Humus"
land_cover$Ground[grepl("humusiert.Acker", land_cover$Art_TXT) | grepl("humusiert.Intensivkultur", land_cover$Art_TXT)] <- "Acre/Orchard"

land_cover$Ground[grepl("bestockt", land_cover$Art_TXT)] <- "Forest"

land_cover$Ground[grepl("befestigt.Strasse", land_cover$Art_TXT) |
                      grepl("befestigt.Verkehr", land_cover$Art_TXT) |
                      grepl("befestigt.uebrige_befestigte.Parkplatz", land_cover$Art_TXT)|                  grepl("humusiert.uebrige_humusierte.Verkehrsteilerflaeche", land_cover$Art_TXT) | grepl("befestigt.Trottoir", land_cover$Art_TXT) | grepl("befestigt.Wasser", land_cover$Art_TXT) | grepl("befestigt.uebrige", land_cover$Art_TXT) | grepl("befestigt.Bahn", land_cover$Art_TXT)] <- "Urban Habitat"

land_cover$Ground[grepl("humusiert.Gartenanlage", land_cover$Art_TXT) | grepl("humusiert.uebrige", land_cover$Art_TXT)] <- "Park"

land_cover$Ground[grepl("humusiert.Gartenanlage.Gartenanlage_Hausumschwung", land_cover$Art_TXT) | grepl("befestigt.uebrige_befestigte.Hausumschwung", land_cover$Art_TXT)] <- "Area around building"

land_cover$Ground[grepl("Gewaesser", land_cover$Art_TXT)] <- "Water"

# check whether there are no grouped land uses
land_cover$Art_TXT[which(is.na(land_cover$Ground))]

unique(land_cover$Ground)
```

```{r}
# summarise grouping of Art_TXT
summary_Art_TXT <- land_cover |> 
  st_drop_geometry() |> 
  group_by(Ground) |> 
  summarise(Art_TXT = paste(unique(Art_TXT), collapse = "; ")) |> 
  arrange(Ground)

# write a new csv file
#write.csv(summary_Art_TXT, "data/summary_Art_TXT.csv")
```

#### Determine Home

```{r}
#| warning: false

# some data points are recorded at home. these have to be marked as "Home" in the land cover layer.

# determine home
home <- data.frame(E=2684988, N=1246066)

# ensure correct CRS
home <- home |>
  st_as_sf(coords = c('E', 'N'), crs=2056)

# set a buffer of 0.1m
home_buffer <- st_buffer(home, dist=0.5)

# subtract the buffer from polygon
clipped_lc <- st_difference(land_cover, home_buffer)

# add the same columns as in land_cover
home_buffer <- home_buffer |> 
  mutate(Art_TXT = NA,
         Ground = "Home")

# bind data set together
land_cover <- rbind(clipped_lc, home_buffer)

rm(home_buffer, clipped_lc)
```

```{r}
# create a vector data set without invalid land cover classes
land_cover_omit <- land_cover |>
  filter(!(Ground == "Water" | Ground == "Building" | Ground == "Home"))
```

```{r}
# create a vector data set without invalid land cover classes
land_cover_omit <- land_cover |>
  filter(!(Ground == "Water" | Ground == "Building"))
```

### GPS Data

#### Read in

```{r}
# list all files to be read with the ending .gpx
# $ = has to be at the end
# function produces a vector of the names of files in the named directory.
files <- list.files(path = "data_tractive", pattern = "\\.gpx$", full.names = TRUE)

# read and append all files
data_list <- lapply(files, function(f) st_read(f, layer = "track_points", quiet = TRUE))

# instead of manually calling rbind(data_list[[1]], data_list[[2]], ...), do.call() automates it.
merged_data <- do.call(rbind, data_list)

rm(files, data_list)
```

```{r}
# select only necessary columns
data <- merged_data[, c("track_seg_point_id", "time", "geometry")]

# WGS 4326, which is used by the package suncalc
data <- st_transform(data, crs = 4326)

# ensure correct time zone (Europe/Zurich)
data$time <- with_tz(data$time, "Europe/Zurich")

# add coordinates as separate columns
data <- cbind(data, st_coordinates(data))

# create a date, weekday and month column
data <- data |>
    dplyr::rename(
    "lon" = X,
    "lat" = Y) |> 
  mutate(date = as.Date(time),
         weekday = weekdays(time),
         month = lubridate::month(time, label=TRUE, abbr=FALSE, locale="EN-uk"))

rm(merged_data)
```

#### Calculate Sunrise and Sunset

Now that the data set still has in the coordinate system which the package «suncalc» uses, we calculate the sunrise and sunset time for each day

```{r}
# calculate sunrise and sunset time for each data point
sun_times <- getSunlightTimes(data = data, keep = c("sunrise", "sunset"), tz = "Europe/Zurich")
data$sunrise <- sun_times$sunrise
data$sunset <- sun_times$sunset

# check whether its day or night
data <- data |>
  mutate(ID = ifelse(time >= sunrise & time <= sunset, "Day", "Night"))

rm(sun_times)
```

#### Correct Coordinate System

```{r}
# geometry in metres, not degrees / ensure correct CRS
data <- st_transform(data, crs=2056)

# add coordinates as separate columns (in metres)
data <- cbind(data, st_coordinates(data))

# rename X and Y
data <- data |>
    dplyr::rename(
    "E" = X,
    "N" = Y)
```

## Invalid Data Points Handling

### Duplicates in time Column

```{r}
# remove the duplicated rows in the time column
data <- data|> 
  distinct(time, .keep_all=TRUE)

# generate unique id
data$track_seg_point_id = seq(1, nrow(data))
```

### Calculate and Investigate Speed

```{r}
# calculate speed
data <- function_speed(data=data, datetime=data$time, geometry=data$geometry)
```

check whether there are points in the same place with speed \> 0 km/h: 0 suspicious data points.

```{r}
data |>
  arrange(track_seg_point_id, time) |>
  mutate(
    speed_kmh = ifelse(speed_kmh < 0.1, 0, speed_kmh),  # treat near-zero speeds as 0
    e_lead = lead(E),
    n_lead = lead(N),
    same_coords = E == e_lead & N == n_lead,
    suspicious = same_coords & speed_kmh != 0
  ) |>
  filter(suspicious)
```

```{r}
# delete data points with speed >= 48 km/h
data <- data |>
  filter(!(speed_kmh >= 48))

# vet data point
data <- data |>
  filter(!(N >= 1246400))

# generate new unique id
data$track_seg_point_id = seq(1, nrow(data))
```

### Check Invalid Data Points (on Water or Building)

```{r}
# join tacking points incl. invalid land cover
data <- st_join(x=data, y=land_cover, join=st_intersects)

# check how many points are on invalid ground cover
nrow(filter(data, Ground == "Water" | Ground == "Building"))

# check how many points are at home
nrow(filter(data, Ground == "Home"))

# as they do not represent home accurately, we remove those 1048 data points
data <- data |>
  filter(!(Ground == "Home"))

# write new geopackage
#st_write(data, "data/data_point_incl.gpkg", layer = "layer", append=FALSE)
```

### Reassign Invalid GPS Points

The reassigning was generated with the help of [ChatGPT](https://chatgpt.com/share/680fa976-3e74-8012-82d3-34a48188277c){target="_blank"}

```{r}
#| warning: false

# determine invalid classes
invalid_classes <- c("Water", "Building", "Home")

#invalid_classes <- c("Water", "Building")

# rename geometry to original_geometry
names(data)[names(data)=="geometry"] <- "original_geometry"
st_geometry(data) <- "original_geometry"

# split GPS points into valid and invalid
invalid <- data |>
  filter(Ground %in% invalid_classes)

valid <- data |>  
  filter(!Ground %in% invalid_classes)

valid_gpkg <- land_cover |>
  filter(!Ground %in% invalid_classes)

# find nearest valid polygon for each invalid GPS point
nearest <- st_nearest_feature(invalid, valid_gpkg)

# get corresponding centroids
# centroid = middle of valid polygon
replacement_points <- st_centroid(valid_gpkg[nearest, ])

# replace geometries
invalid$geometry <- st_geometry(replacement_points)

# calculate distance between original and new locations
invalid$correction_distance <- st_distance(
  invalid$original_geometry,
  invalid$geometry,
  by_element=TRUE
)

# create new columns in data_point
valid$geometry <- valid$original_geometry

# add 0, if class was already valid
valid$correction_distance <- 0

# join the two data frames
data_reassigned <- rbind(valid, invalid)

# sort according to column time
data_reassigned <- data_reassigned[order(data_reassigned$time),]

# reset index
row.names(data_reassigned) <- NULL

# reset geometry
st_geometry(data_reassigned) <- "geometry"

# join with new points
data_reassigned <- data_reassigned |> 
  dplyr::select(-c(Art_TXT, Ground))

data_reassigned = st_join(data_reassigned, land_cover, join=st_intersects)

# rename geometry column again
names(data)[names(data)=="original_geometry"] <- "geometry"

rm(valid, invalid, valid_gpkg, invalid_classes, nearest, replacement_points)
```

```{r}
# check whether there are still invalid classes in the data set
# with st_centroid() it can happen that the new point is not in the desired polygon. But as it is computationally less intensive than other methods we decided to stick with that.
sum(data_reassigned$Ground=="Water")
sum(data_reassigned$Ground=="Building")
sum(data_reassigned$Ground=="Home")

# because the numbers are so low, the data points which are still on invalid land cover classes are deleted
data_reassigned <- data_reassigned |>
  filter(!(Ground=="Water" | Ground=="Building" | Ground=="Home"))

# data_reassigned <- data_reassigned |> 
#   filter(!(Ground=="Water" | Ground=="Building"))

# check if only the desired land cover classes are remaining in the new data set
unique(data_reassigned$Ground)
unique(data$Ground)

# write new geopackage
st_write(data_reassigned, "data/data_point_incl_re.gpkg", layer = "layer", append=FALSE)
```

## Day and Night Data Sets

```{r}
# separate by day/night
data_day_re <- data_reassigned |> 
  filter(ID == "Day")

data_night_re <- data_reassigned |> 
  filter(ID == "Night")
```

## Available Land Cover

Set home and calculate a buffer of a radius = 350 m to represent available land cover around home range (as in Thomas et al., 2014).

```{r}
home_350 <- st_buffer(home, dist=350)

home_350 <- home_350 |> 
  mutate(month = NA,
         method = "Available Land Cover",
         ID = "Available Land Cover",
         area_ha = as.numeric(st_area(geometry))/10000)

rm(home)
```

## Territory Estimation (KDE & MCP)

```{r}
#| warning: false

# as there was an error about geometry, we reset it before calculating the territory
# reset geometry
st_geometry(data) <- "geometry"

# for data set incl invalid data points
data_mcp_incl <- function_mcp(data=data, ID="incl.")
data_kde_incl <- function_kde(data=data, ID="incl.")
```

```{r}
#| warning: false

# for data set with reassigned data points
# reset geometry
st_geometry(data_reassigned) <- "geometry"

data_mcp_re <- function_mcp(data=data_reassigned, ID = "reassigned")
data_kde_re <- function_kde(data=data_reassigned, ID = "reassigned")
```

```{r}
#| warning: false

# reset geometry
st_geometry(data_day_re) <- "geometry"
st_geometry(data_night_re) <- "geometry"

data_day_mcp_re <- function_mcp(data=data_day_re, ID = "Day")
data_night_mcp_re <- function_mcp(data=data_night_re, ID = "Night")
data_day_kde_re <- function_kde(data=data_day_re, ID = "Day")
data_night_kde_re <- function_kde(data=data_night_re, ID = "Night")

rm(data_day_re, data_night_re)
```

```{r}
# estimated territory contains the information on the area of territory per month and separated by estimation method (KDE & MCP). it also contains the information on the home buffer (available land cover)
estimated_territory <- rbind(data_mcp_incl, data_kde_incl, data_mcp_re, data_kde_re, data_day_mcp_re, data_night_mcp_re, data_day_kde_re, data_night_kde_re, home_350)

# write new geopackage
#st_write(estimated_territory, "data/estimated_territory.gpkg", layer = "layer", append=FALSE)

rm(estimated_territory)
```

## Land Cover and Territory

```{r}
#| warning: false

# in the buffer there is no weighting / no preference of land cover, just represents what's there
data_buffer_grouped <- function_group(data=home_350, data_land_cover=land_cover_omit, method="buffer", method_area_perc="buffer_perc", buffer=TRUE)

# group by Ground and calculate the preferred area per Ground in % according to MCP for the data set incl. invalid data points.
# as there was again a problem with the geometry column, we remove it here again (even though it is removed in the function), just to be sure.
data_mcp_grouped_incl <- function_group(data=data_mcp_incl, data_land_cover=land_cover, method="mcp_incl", method_area_perc="mcp_incl_perc")
data_mcp_grouped_incl <- subset(data_mcp_grouped_incl, select=-c(geometry))

# group by Ground and calculate the preferred area per Ground in % according to KDE  for the data set incl. invalid data points.
data_kde_grouped_incl <- function_group(data=data_kde_incl, data_land_cover=land_cover, method="kde_incl", method_area_perc="kde_incl_perc")
data_kde_grouped_incl <- subset(data_kde_grouped_incl, select=-c(geometry))

# group by Ground and calculate the preferred area per Ground in % according to MCP for the data set with reassigned invalid data points.
data_mcp_grouped_re <- function_group(data=data_mcp_re, data_land_cover=land_cover_omit, method="mcp_re", method_area_perc="mcp_re_perc")
data_mcp_grouped_re <- subset(data_mcp_grouped_re, select=-c(geometry))

# group by Ground and calculate the preferred area per Ground in % according to KDE for the data set with reassigned invalid data points.
data_kde_grouped_re <- function_group(data=data_kde_re, data_land_cover=land_cover_omit, method="kde_re", method_area_perc="kde_re_perc")
data_kde_grouped_re <- subset(data_kde_grouped_re, select=-c(geometry))

# group by Ground and calculate the preferred area per Ground in % according to MCP for the data set with reassigned invalid data points for daytime.
data_day_mcp_grouped_re <- function_group(data=data_day_mcp_re, data_land_cover=land_cover_omit, method_area="mcp_day", method_area_perc="mcp_day_perc")
data_day_mcp_grouped_re <- subset(data_day_mcp_grouped_re, select=-c(geometry))

# group by Ground and calculate the preferred area per Ground in % according to MCP for the data set with reassigned invalid data points for nighttime.
data_night_mcp_grouped_re <- function_group(data=data_night_mcp_re, data_land_cover=land_cover_omit, method_area="mcp_night", method_area_perc="mcp_night_perc")
data_night_mcp_grouped_re <- subset(data_night_mcp_grouped_re, select=-c(geometry))

# group by Ground and calculate the preferred area per Ground in % according to KDE for the data set with reassigned invalid data points for daytime.
data_day_kde_grouped_re <- function_group(data=data_day_kde_re, data_land_cover=land_cover_omit, method_area="kde_day", method_area_perc="kde_day_perc")
data_day_kde_grouped_re <- subset(data_day_kde_grouped_re, select=-c(geometry))

# group by Ground and calculate the preferred area per Ground in % according to KDE for the data set with reassigned invalid data points for nighttime.
data_night_kde_grouped_re <- function_group(data=data_night_kde_re, data_land_cover=land_cover_omit, method_area="kde_night", method_area_perc="kde_night_perc")
data_night_kde_grouped_re <- subset(data_night_kde_grouped_re, select=-c(geometry))

rm(data_mcp_incl, data_kde_incl, data_mcp_re, data_kde_re, data_day_mcp_re, data_night_mcp_re, data_day_kde_re, data_night_kde_re)
```

```{r}
# create a new data frame combining all the beforehand generated data frames
list_df = list(data_mcp_grouped_re, data_kde_grouped_re, data_day_mcp_grouped_re, data_night_mcp_grouped_re, data_day_kde_grouped_re, data_night_kde_grouped_re)

data_joined_re <- list_df |> 
  reduce(inner_join, by=c("month", "Ground"))

data_joined_re <- inner_join(data_joined_re, data_buffer_grouped, by="Ground")
```

```{r}
# create long format
data_joined_long <- data_joined_re |> 
  dplyr::select(-buffer_perc, -mcp_re_perc, -kde_re_perc, -mcp_day_perc, -kde_day_perc, -mcp_night_perc, -kde_night_perc) |> 
  tidyr::pivot_longer(
    cols = -c(month, Ground),
    names_to = "Method",
    values_to = "Area"
  ) |> 
  separate(Method, c("Method", "ID"), "_",  fill="right")

# write new csv
#write.csv(data_joined_long, "data/data_lc_territory_long.csv")
```

```{r}
#| warning: false

# calculate selection ratio of land cover
# this should tell us, which land cover the cat prefers and avoids compared to available land cover
data_joined_re <- data_joined_re |>
  mutate(selection_ratio_mcp_re=mcp_re_perc/buffer_perc,
  log_selection_mcp_re=log(selection_ratio_mcp_re),
  selection_ratio_kde_re=kde_re_perc/buffer_perc,
  log_selection_kde_re=log(selection_ratio_kde_re))

# write new csv 
#write.csv(data_joined_re, "data/data_lc_territory_re.csv")

rm(data_joined_long_dn, data_joined_re, list_df, data_mcp_grouped_incl, data_kde_grouped_incl, data_mcp_grouped_re, data_kde_grouped_re, data_day_mcp_grouped_re, data_night_mcp_grouped_re, data_day_kde_grouped_re, data_night_kde_grouped_re, data_buffer_grouped)
```

## Monte Carlo Simulation for one Month with 50 Iterations

### Data Preparation

Only choose one month from `data` - we choose June because there are the most data points.

```{r}
##reset geometry column
st_geometry(data) <- "geometry"

##select June
data_june <- data |> 
  filter(month=="June")

##select necessary columns
data_june <- data_june[ , c("track_seg_point_id", "E", "N", "geometry")]
```

Generate normally distributed values.

```{r}
# set number of iterations
iterations <- 50

# initialise empty matrices for new x and y
x_mat = matrix(NA, nrow(data_june), iterations) 
y_mat = matrix(NA, nrow(data_june), iterations)

# Fill empty matrices with normally distributed values. According to the GPS collar provider, the accuracy is 8m. We used 8/3 for the standard deviation as 3*sd usually contains around 99.7% of all values.  
set.seed(111)

for(rows in 1:nrow(data_june)){
  for(columns in 1:iterations){
    x_mat[rows,columns] <- rnorm(1,0,8/3)
    y_mat[rows,columns] <- rnorm(1,0,8/3)
  }
}

# add normally distributed values to coordinates and store it in x_mat and y_mat respectively
for(rows in 1:nrow(data_june)){
  for(columns in 1:iterations){
    x_mat[rows,columns] <- x_mat[rows,columns]+data_june$E[rows]
    y_mat[rows,columns] <- y_mat[rows,columns]+data_june$N[rows]
  }
}

# initiate empty data frame for storing the combined x- and y-coordinates obtained from the iterations
geom_iterations <- data.frame(matrix(ncol=0, nrow=nrow(data_june)))

# fill the empty data frame
for(columns in 1:iterations){
  xy_mat <- cbind(x_mat[ ,columns], y_mat[ ,columns])
  xy_sf <- st_as_sf(data.frame(xy_mat), coords=c(1,2), crs=2056)
  geom_iterations[ , columns] <- xy_sf
  names(geom_iterations)[columns] <- paste("Iteration", columns)
}

# remove some now unnecessary dataframes etc.
#rm(x_mat, y_mat, xy_mat, xy_sf)

# combine points from iterations with the original data
data_june_all <- cbind(data_june$geometry, geom_iterations) 
```

```{r}
# rename geometry column in `data_june`
data_june_all <- data_june_all |>
  rename("Original" = geometry)

# transform to long
june_all_long <- data_june_all |> 
  data.frame() |> 
  pivot_longer(
    cols=everything(),
    names_to="iteration",
    values_to="geometry"
  )

# transform to sf-object
june_all_long <- st_as_sf(june_all_long)

# join with land cover data
land_cover_reduced <- land_cover[ ,c("geometry", "Ground")]
june_all_ground <- st_join(x=june_all_long, y=land_cover_reduced, join=st_intersects)

# calculate percentage of different grounds, drop geometry & transform to data frame
june_all_ground_perc <- june_all_ground |> 
  dplyr::group_by(iteration, Ground) |> 
  summarise(number_points=n()) |> 
  mutate(percentage=round((number_points/sum(number_points)*100), digits=2))|>
  st_drop_geometry() |> 
  data.frame()
```

```{r}
# transform to wide (first remove percentage)
perc_wide <- june_all_ground_perc |>
  dplyr::select(-percentage) |> 
  pivot_wider(names_from=Ground, values_from=number_points) |> 
  data.frame()

rownames(perc_wide) <- perc_wide$iteration #give rows a name

perc_wide <- perc_wide |> 
  dplyr::select(-Home, -iteration) #remove unneccessary columns
```

### Chi-Squared-Test

```{r}
#| warning: false

# save as two different data sets & add additional column for p-value in the iterations data frame
perc_wide_orig <- perc_wide[which(rownames(perc_wide) == "Original"), ]

perc_wide_iter <- perc_wide[which(rownames(perc_wide) != "Original"), ]
perc_wide_iter$"p-value from chi-squared-test" <-  NA 

# run loop for chi-square-testing the original against each iteration
for (rows in 1:nrow(perc_wide_iter)){
  vec_orig <- as.numeric(perc_wide_orig)
  vec_iter <- as.numeric(perc_wide_iter[rows, 1:7 ])
  chisq_test <- chisq.test(vec_orig, vec_iter)
  perc_wide_iter$`p-value from chi-squared-test` <- chisq_test$p.value
}

#write.csv(perc_wide_orig, "data/perc_wide_orig.csv")
#write.csv(perc_wide_iter, "data/perc_wide_iter.csv")
```

```{r}
#| warning: false

# also perform chi-squared-test on the total of the iterations
# 1st initiate empty data frame
perc_significance_total <- data.frame(matrix(ncol=ncol(perc_wide), nrow=3))
rownames(perc_significance_total) <- c("mean of iterations", "median of iterations", "sum of iterations")
colnames(perc_significance_total) <- colnames(perc_wide)

# insert mean, median and sum from all iterations of all land cover types
perc_significance_total[1, ] <- as.numeric(colMeans(perc_wide_iter))
perc_significance_total[2, ] <- as.numeric(apply(perc_wide_iter,2,median))
perc_significance_total[3, ] <- as.numeric(colSums(perc_wide_iter))

# add additional column for p-value
perc_significance_total$"p-value from chi-squared-test" <-  NA #add additional column for p-value

# run loop for mean, median and sum
for (rows in 1:nrow(perc_significance_total)){
  vec_orig <- as.numeric(perc_wide_orig)
  vec_iter <- as.numeric(perc_significance_total[rows,1:7 ])
  chisq_test <- chisq.test(vec_orig, vec_iter)
  perc_significance_total$`p-value from chi-squared-test` <- chisq_test$p.value
}

#write.csv(perc_significance_total, "data/perc_significance_total.csv")
```
