---
title: "Preprocessing: Investigating Cat Movement Patterns, Space Use, and Environmental Influences"
author: "Selina Lepori"
date: 2025-07-04
format:
  html:
    toc: true
    css: styles.css
    code-fold: false
    theme:
      dark: darkly
execute:
  warning: false
  message: false
lang: en
lightbox: true
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| include: false

# sets background of figures = transparent (otherwise white borders around ggplots)
knitr::opts_chunk$set(dev.args = list(bg = "transparent"))
```

# Load Libraries and Functions

```{r}
#| include: false

rm(list = ls())
```

```{r}
#| warning: false

library(sf)
library(sp)
library(lubridate)
library(tidyverse)
library(quarto)
library(data.table)
library(suncalc)
library(ggdark)
library(tmap)
library(kableExtra)
library(terra)
library(smoothr)
library(adehabitatHR)

source("speed.R")
source("summary_table.R")

create_files = FALSE

```

# Read, Tidy, and Join Data Frames

## Determine Home

At home, the tracker connects with the WiFi and should be in a power saving zone and stop sending data. However, it still sends data irregularly. Therefore, the home coordinates have to be determined and a buffer of 8 m radius has to be added. This data frame can be used later to remove the wrong GPS data points and the land cover data within this buffer.

```{r}
#| include: false

# determine home
home <- data.frame(name=c("home"), E=c(2684990), N=c(1246068))
```

```{r}
# ensure correct CRS
home <- home |>
  st_as_sf(coords = c('E', 'N'), crs=2056)

# set a buffer of 8 m
home_buffer <- st_buffer(home[,2], dist=8)
```

## Meteo Data

```{r}
# meteo data
data_meteo <- read.csv("data/sma-meteoswiss-dataset.csv")

# select only precipitation column and ensure correct time zone
data_meteo <- data_meteo |> 
  dplyr::select(time, rain10minsum) |> 
  na.omit() |> 
  mutate(time = as.POSIXct(time, format = "%d/%m/%Y %H:%M", tz = "Europe/Zurich"))
```

## GPS

### Read in

```{r}
# list all files to be read with the ending .gpx
# $ = has to be at the end
# function produces a vector of the names of files in the named directory.
files <- list.files(path = "data_tractive", pattern = "\\.gpx$", full.names = TRUE)

# read and append all files
data_list <- lapply(files, function(f) st_read(f, layer = "track_points", quiet = TRUE))

# instead of manually calling rbind(data_list[[1]], data_list[[2]], ...), do.call() automates it.
merged_data <- do.call(rbind, data_list)

rm(files, data_list)
```

```{r}
# select only necessary columns
data <- merged_data[, c("track_seg_point_id", "time", "geometry")]

# WGS 4326, which is used by the package suncalc
data <- st_transform(data, crs = 4326)

# ensure correct time zone (Europe/Zurich)
data$time <- with_tz(data$time, "Europe/Zurich")

# add coordinates as separate columns
data <- cbind(data, st_coordinates(data))

# create a date, weekday and month column
data <- data |>
    dplyr::rename(
    "lon" = X,
    "lat" = Y) |> 
  mutate(date = as.Date(time),
         weekday = weekdays(time),
         month = lubridate::month(time, label=TRUE, abbr=FALSE, locale="EN-uk"))

rm(merged_data)
```

### Correct Coordinate System

```{r}
# geometry in metres, not degrees / ensure correct CRS
data <- st_transform(data, crs=2056)

# add coordinates as separate columns (in metres)
data <- cbind(data, st_coordinates(data))

# rename X and Y
data <- data |>
    dplyr::rename(
    "E" = X,
    "N" = Y)
```

### Remove invalid Data Points

```{r}
# there is one data point recorded at the vet which must be deleted
data <- data |>
  filter(!(N >= 1246400))
```

```{r}
# remove the duplicated rows in the time column
data <- data|> 
  distinct(time, .keep_all=TRUE)

# generate unique id
data$track_seg_point_id = seq(1, nrow(data))
```

```{r}
#| warning: false

# select necessary columns
# data <- data |>
#   dplyr::select(time, date, month, sunrise, sunset, ID, geometry, E, N, lon, lat)

# set geometry column
st_geometry(data) <- "geometry"

# ensure correct time zone
data$time = as.POSIXct(data$time, format = "%Y-%m-%d %H:%M:%S", tz = "Europe/Zurich")
```

As mentioned above, the GPS tracking points which are located within the buffer representing home shall be omitted.

```{r}
# get GPS tracking points within the home buffer
home_points <- st_intersection(st_as_sf(data), home_buffer)

# setdiff(x, y) finds all rows in x that aren't in y.
# so removes every row from data which is in home_points
data <- setdiff(data, home_points)
```

```{r}
#| include: FALSE

if (create_files) {
  st_write(data, "data/data_raw.gpkg", layer = "layer", append=FALSE)
}
```

```{r}
#| include: FALSE

rm(home_buffer, home, home_points)
```

### Calculate Time Lag and Check Speed Value

Speed is calculated for indicating invalid data points.

```{r}
# calculate timelag and speed
data <- function_timelag(data=data, datetime=data$time, geometry=data$geometry)

data <- function_speed(data=data, datetime=data$time)
```

Speed larger than 48 km/h is not possible, therefore these points will be omitted.

```{r}
# delete data points with speed >= 48 km/h
data <- data |>
  filter(!(speed_kmh >= 48))

# generate new unique id
data$track_seg_point_id = seq(1, nrow(data))
```

The median interval is around 10 minutes, as was to be expected. The mean interval is larger for every month, which can be explained due to larger time gaps when, for example, the cat was at home or wasn't wearing the tracker.
After the deletion of intervals which are smaller than 9 minutes, the mean value increases but the median value for the intervals remain more or less the same. The removal of intervals smaller than 9 minutes was conducted to simplify the calculation of static and non-static phases, which will be described in the next part of the project.

```{r}
#| label: tbl-overview
#| warning: false
#| tbl-cap: Overview of time lags in the gps points data set.
#| echo: false

table <- create_summary_table(data)

knitr::kable(table)
```

```{r}
#| include: false

rm(table)
```

```{r}
# function for calculating timelag
data <- function_timelag(data = data, datetime = data$time, geometry = data$geometry)

data <- data |> 
 filter(!is.na(timelag), timelag > 540)
```

```{r}
# recaclulate timelag and speed
data  <- function_timelag(data = data, datetime = data$time, geometry = data2$geometry)

data <- data |> 
 filter(!is.na(timelag))

data <- function_speed(data=data, datetime=data$time)

# create incrementing ID
data$point_id = seq(1, nrow(data))
```

Omitting the points with an interval smaller 9 minutes simplified the speed calculation as well. The highest speed value is around 0.60 m/s or 2.17 km/h, respectively, which is plausible.

```{r}
#| label: tbl-speed
#| tbl-cap: This table shows the top 10 speed values after cleaning the data.
#| warning: false
#| echo: false

table <- data |>
  st_drop_geometry() |>
  dplyr::select(time, speed_ms, speed_kmh) |>
  arrange(desc(round(speed_ms, digits=2))) |>
  head(10)

knitr::kable(table)
```

```{r}
#| include: false

rm(table)
```

# Section 1: Within Home Range Structure and Precipitation Effects

## Precipitation

### Determine Rainy Weather

Short periods (< 1h) without any rain should be combined with the rain period before (and after, if there is an after period).

Description of the code:

1. create a new column which tells us whether it's raining (TRUE) or not (FALSE).

2. create a new data frame («rle_data_meteo») which summarises the length of each (TRUE/FALSE) sequence incl. ID for each sequence.

3. Loop: Loop through «rle_data_meteo» and look for dry periods (value = FALSE), which are shorter than one hour (length < 6). When such a point is found, check whether the point before belongs to a rainy period. If that is the case, merge the current point ID with the ID from the point before and determine the current point as rainy (value = TRUE). Furthermore, check whether the following point is a rainy point. If so, merge the IDs again, so that the subsequent period belongs to the current period as well. In this case, there is no need to change the value from FALSE to TRUE, as it is already TRUE.

4. Create a new column «id» in the meteo data and add the id determined before to the data frame. «rep» replicates the x values by the times value. Do the same for precipitation.

5. In the end, redefine short precipitation events (<= 10 minutes) to FALSE.

```{mermaid}
flowchart LR
  A[start loop] --> B{short dry period?}
  B -- yes --> C{rain before?}
  C -- yes --> D(adapt id, rain=TRUE)
  D --> E{rain after?}
  E -- yes --> F[adapt id]
  B -- no --> A
  C -- no --> A
  E -- no --> A
```

This [flow chart](https://quarto.org/docs/authoring/diagrams.html){target="_blank"} describes the following loop which determines rainy conditions.

source for the code: [ChatGPT](https://chatgpt.com/share/682ae22a-2e18-8012-99b7-abda090599a8){target="_blank"}

```{r}
# create a new column which tells us whether it's raining (TRUE) or not (FALSE)
data_meteo <- data_meteo |> 
  na.omit() |> 
  mutate(precipitation_org = rain10minsum > 0)

# create a run-length ID for grouping (no) rain
# create a new data frame which summarises the length of each (TRUE/FALSE) sequence
# https://www.r-bloggers.com/2009/09/r-function-of-the-day-rle-2/
rle_data_meteo <- rle(data_meteo$precipitation_org)
rle_data_meteo <- data.frame(
  length = rle_data_meteo$lengths, # length of sequence
  value = rle_data_meteo$values, # TRUE or FALSE
  id = seq_along(rle_data_meteo$lengths) # create unique ID for each (no) rain period
)

# combine short dry periods (< 6 time steps = 60 minutes) with rain periods
for (i in 1:nrow(rle_data_meteo)) {
  # if precipitation == FALSE and length < 6
  # && compares only the first element of each vector
  if (!rle_data_meteo$value[i] && rle_data_meteo$length[i] < 6) {
    # merge short dry period to previous and/or next rain period
    # if obs - 1 is raining and current short period is not
    if (i > 1 && rle_data_meteo$value[i - 1]) {
      # id of current == id of obs - 1
      rle_data_meteo$id[i] <- rle_data_meteo$id[i - 1]
      rle_data_meteo$value[i] <- TRUE # change to rainy period
    }
    # if obs + 1 is raining and current short period is not
    if (i < nrow(rle_data_meteo) && rle_data_meteo$value[i + 1]) {
      # id of obs + 1 == id of current
      rle_data_meteo$id[i + 1] <- rle_data_meteo$id[i]
    }
  }
}

# create new group IDs and precipitation values
# so that the number of observations is equal to the original data_meteo
data_meteo$id <- rep(x = rle_data_meteo$id, times = rle_data_meteo$length)
data_meteo$precipitation <- rep(rle_data_meteo$value, rle_data_meteo$length)

# redefine short precipitation events (<= 10 minutes) to FALSE
data_meteo <- data_meteo |> 
  group_by(id) |> 
  mutate(
    precipitation = ifelse(precipitation & n() == 1, FALSE, precipitation)) |> 
    # if precipitation == TRUE but for 10 minutes only -> FALSE 
  ungroup()
```

```{r}
#| include: FALSE

rm(rle_data_meteo)
```

Join with weather data. As the time stamps are not exactly the same in the GPS dataset and the meteo dataset, the data points have to be joined with the nearest match on the column «time».

```{r}
# data frame to data table for joining
setDT(data)
setDT(data_meteo)

# join with nearest match
# right join
data <- data_meteo[data, roll = "nearest", on = "time"]

# remove some columns
data <- data |>
  dplyr::select(-c(rain10minsum, precipitation_org, id))
```

## Phases of the Day

### Determine Phases of the Day (Night, Dawn, Daytime, Dusk)

The getSunlightTimes function of the library «suncalc» is a nice function to determine different phases of the day. With this function I determine dawn, daytime, dusk, and nighttime.

```{r}
data

# get sunlight times
# suntimes are correct, even with daylight saving time (checked online)
sun_times <- getSunlightTimes(data = data, tz = "Europe/Zurich")

# https://stackoverflow.com/questions/48038236/using-dplyr-mutate-with-conditions-based-on-multiple-columns
data <- data |>
  mutate(dayphase = case_when(
    time >= nauticalDawn & time <= sunriseEnd ~ "Dawn",
    time >= sunsetStart & time <= nauticalDusk ~ "Dusk",
    time >= sunriseEnd & time <= sunsetStart ~ "Daytime",
    time < nauticalDawn | time > nauticalDusk ~ "Nighttime"
  ))

# keep only selected new columns
data <-  data |> 
  dplyr::select(-c(names(sun_times[4:17]))
)

# check whether each gps tracking point is grouped
unique(data$dayphase)
```

```{r}
#| include: FALSE

rm(sun_times)
```

# Section 2: Movement Analysis

According to [Tractive](https://help.tractive.com/hc/en-us/articles/115002612125-Why-is-my-tracker-position-inaccurate){target="_blank"}, the tracking points' positional inaccuracy is around 8 m. This is why I stick with this number to determine stationary points. The idea is to check whether the following point (i+1) is within the buffer of the first point (i). If that's the case, the next point (i+2) is compared to the original first (i) point as well and so on. In this way, stationary phases are determined. As soon as a point is not within the buffer of the first point, it is declared as non-static. The «else if» part of the loop ensures, that the first point of the static segment is declared as static as well.

```{r}
data <- st_as_sf(x = data, crs = 2056)

# set a buffer of 8 m around each gps tracking point
geom_buffer = st_buffer(data, dist=8)

# select column geometry only and rename as geom_buffer
geom_buffer <- geom_buffer |>
  dplyr::select(geometry) |>
  dplyr::rename("geom_buffer" = geometry)

# add the buffer geometry to the original data
data$geom_buffer = st_geometry(geom_buffer)
```

```{r}
#| include: false

rm(geom_buffer)
```

```{mermaid}
flowchart LR
  A[start loop & id=1] --> B{current ∩ original?}
  B -- yes --> C["static"]
  B -- no --> D{current ∩ next?}
  D -- yes --> E["static" & id=current]
  D -- no --> F["not static" & id=current]
```

This [flow chart](https://quarto.org/docs/authoring/diagrams.html){target="_blank"} describes the following code, which determines whether a point and its buffer is static or non-static. ∩ = intersection.

```{r}
# create a new column to differ stationary from moving phases
data$static = NA

# set id to 1 and change after steps
id = 1

# loop through the data set and determine whether the following buffer is intersecting with the buffer of the original point
# if so: static = TRUE. repeat until following buffer isn't intersecting buffer.
# i > 1 ensures, that the very first data point is not declared as static.
for (i in 1:nrow(data)) {
  if (i > 1 && length(st_intersection(data$geom_buffer[i], data$geom_buffer[id])) > 0) {
    data$static[i] = "Static"
  }

# also check whether the following point is static and if so, declare the point before as static.
# otherwise the first point of the static segment will be declared as non-static.
# i > 1 ensures, that the very first data point is not declared as static.
  else if (i > 1 && length(st_intersection(data$geom_buffer[i], data$geom_buffer[i+1])) > 0) {
    data$static[i] = "Static"
    id = i
  }

# if both cases before are false, declare as non-static.
  else {
    data$static[i] = "Not Static"
    id = i
  }
}
```

## Plot Sample of (Non-)Static Data

Create a small sample of the data points to check the method visually.

```{r}
# data_sample_plot <- data |>
#     st_as_sf(coords = c("E", "N"), crs = 2056, remove = FALSE) |>
#     filter(
#       time >= as.POSIXct("2024-04-01 00:00:00", tz = "Europe/Zurich"), 
#       time < as.POSIXct("2024-04-01 18:00:00", tz = "Europe/Zurich")
#       )
```

```{r}
# set a buffer of 8 m around each gps tracking point
# geom_buffer = st_buffer(data_sample_plot, dist=8)
# 
# # select column geometry only and rename as geom_buffer
# geom_buffer <- geom_buffer |>
#   dplyr::select(geometry) |> 
#   dplyr::rename("geom_buffer" = geometry)
```

```{r}
# create a line between the points which takes into consideration the time stamp
# 
# data_line <- data_sample_plot |> 
#   # dissolve to a MULTIPOINT:
#   summarise(do_union = FALSE) |> 
#   st_cast("LINESTRING")
```

```{r}
#| label: fig-static-non-static-sample
#| fig-cap: This sample plot shows clearly which points are declared as static and non-static, according to the method described above.
#| warning: false
#| echo: false

# tmap_mode("plot")
# tm_shape(data_line) +
#   tm_lines() +
#   tm_shape(geom_buffer) +
#   tm_fill(col = "blue", alpha=0.5)  +
#   tm_shape(data_sample_plot) +
#   tm_dots(col = "static", palette = c("black", "red")) +
#   tm_text("point_id") +
#   tm_basemap("OpenStreetMap")
```

```{r}
#| include: false

#rm(geom_buffer, data_line, data_sample_plot)
```






